### 2017-05-30 doc/categories.tex

Models that utilize the sequential structure of language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. In principle, this may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent {\sc Visual} and {\sc Textual}
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementaly build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, are sensitive by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables: 
\begin{enumerate}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{enumerate}

\noindent We split the 5-thousand-image portion of MSCOCO validation data into two parts, fit the 
models on the first part, and compute the 
proportion of variance explained on the second part. 
For comparison we also use the {\sc Sum} model which composes word
embeddings via summation, and uses the same loss function as {\sc
  Visual}. This model is unable to encode information about word
order, and thus is a good baseline here as we investigate the
sensitivity of the networks to positional and structural cues.

\begin{table}
  \centering
  \caption{Proportion of variance in omission scores explained by
    linear regression.}
    \begin{tabular}{l|rrrr}
               & word   & +pos  & +dep  & full \\\hline
     sum       & 0.654  & 0.661 & 0.670 & 0.670 \\
     LM        & 0.358  & 0.586 & 0.415 & 0.601 \\
     textual   & 0.364  & 0.703 & 0.451 & 0.715 \\
     visual    & 0.490  & 0.506 & 0.515 & 0.523 \\
    \end{tabular}
    \label{tab:lr-r2}
\end{table}

Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
Figure~\ref{fig:rsquared} offers a different view of the data, showing
the increase or decrease in $R^2$ for the models relative to {\sc
  LR~+pos}.  
The results reveal that adding additional information on
top of lexical features in case of {\sc Sum} does increase the
explained variance slightly, which is most likely due to the unseen words
in the held out set. More interestingly, there is a sizeable
increase in $R^2$ between {\sc LR~+pos} and {\sc LR~full} in
case of {\sc Visual}, suggesting that 
the omission scores in case of {\sc Visual} depend on the words'
grammatical function in sentences, {\it even after controling for word
identity and linear position.}

@@

In order to find out some of the specific syntactic configurations leading to
increase in $R^2$ between the {\sc LR~word} and {\sc LR~dep} predictors
in case of {\sc Visual}, we next considered all word types with
occurrence counts of at least 100 and ranked them according to how much
better, on average, {\sc LR~+dep} predicted their omission scores 
compared to {\sc LM~word}.

Figure~\ref{fig:top_words} shows the per-dependency 
omission-score distributions for seven top ranked words.
There are clear and large differences in how these words
impact the network's representation depending on what grammatical
function they fulfill. They all have large omission scores when they
occur as {\sc nsubj} (nominal subject) or {\sc root}, likely due to the fact that these
grammatical functions typically have a large contribution to the
complete meaning of a sentence.  Conversely, all have small omission 
scores when appearing as {\sc conj} (conjunct): this is probably because in this position
they share their contribution with the first, often more important,
member of the conjunction, for example: {\it A cow and its baby eating
  grass}. 
% The pattern for {\sc nn} (nominal modifier) is a bit more complicated: for
% four of the words shown (as well as for most other words not shown in
% the figure), the score is very low in this grammatical function--presumably 
% because most words contribute less to the sentence meaning when used  
% as modifiers than as heads (e.g.\ {\it a clock tower}). However,
% for the words {\it zebra} and {\it water}, omission scores are high
% when they act as a nominal modifier {\sc nn}. This appears due to two reasons:
% \begin{enumerate}
% \item For {\it zebra}, there are frequent erroneous parses such as {\it
%     zebra/{\sc nn} browsing/{\sc root}} instead of {\it zebra/{\sc nsubj} browsing/{\sc
%     root}}. The network does not make this mistake, and treats these
%     occurrences of {\it zebra} according to its importance as {\sc nsubj}.
% \item {\it Water} as a modifier often changes the meaning of its head in
%  a visually salient way: e.g.\ {\it water fall, water balloon, water scene, water skiing}, and
%   thus the network learns that this particular word is important in the modifier position.
% \end{enumerate}

@@

As observed in Section \ref{sec:beyondlexical}, 
adding extra information about the position of words 
explains more of the variance in case of {\sc Visual} and especially
{\sc Textual} and {\sc LM}.
Figure \ref{fig:posrqs} shows the coefficients corresponding to the
position variables in {\sc LR~full}. Since the omission scores 
are measured at the end-of-sentence token, the expectation is that 
for {\sc Textual} and {\sc LM}, as language models,  
the words appearing closer to the end of the sentence would have a
stronger effect on the omission scores. This seems to be confirmed by
the plot as the coefficients for these two networks up until the
\emph{antepenult} are all negative.

>>>>

Models that utilize the sequential structure of language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. In principle, this may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent {\sc Visual} and {\sc Textual}
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementally build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, on the other hand, are sensitive by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables: 
\begin{itemize}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{itemize}

\noindent We split the 5000-image portion of MSCOCO validation data into two equal parts, fit the 
models on the first part, and compute the 
proportion of variance explained on the second part. 
For comparison we also use the {\sc Sum} model which composes word
embeddings via summation, and uses the same loss function as {\sc
  Visual}. This model is unable to encode information about word
order, and thus is a good baseline here as we investigate the
sensitivity of the networks to positional and structural cues.

\begin{table}
  \centering
  \caption{Proportion of variance in omission scores explained by
    linear regression.}
    \begin{tabular}{l|rrrr}
               & word   & +pos  & +dep  & full \\\hline
     {\sc Sum}       & 0.654  & 0.661 & 0.670 & 0.670 \\
     {\sc LM}        & 0.358  & 0.586 & 0.415 & 0.601 \\
     {\sc Textual}   & 0.364  & 0.703 & 0.451 & 0.715 \\
     {\sc Visual}    & 0.490  & 0.506 & 0.515 & 0.523 \\
    \end{tabular}
    \label{tab:lr-r2}
\end{table}

\todo[inline]{I think we can improve the discuss of results below.}
Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
Figure~\ref{fig:rsquared} offers a different view of the data, showing
the increase or decrease in $R^2$ for the models relative to {\sc
  LR~+pos}.  
The results reveal that adding additional information on
top of lexical features in the case of {\sc Sum} does increase the
explained variance slightly, which is most likely due to the unseen words
in the held out set. More interestingly, there is a sizeable
increase in $R^2$ between {\sc LR~+pos} and {\sc LR~full} in
the case of {\sc Visual}, suggesting that 
the omission scores in the case of {\sc Visual} depend on the words'
grammatical function in sentences, {\it even after controlling for word
identity and linear position.}

@@

In order to find out some of the specific syntactic configurations leading to
an increase in $R^2$ between the {\sc LR~word} and {\sc LR~+dep} predictors
in the case of {\sc Visual}, we next considered all word types with
occurrence counts of at least 100 and ranked them according to how much
better, on average, {\sc LR~+dep} predicted their omission scores 
compared to {\sc LM~word}.

Figure~\ref{fig:top_words} shows the per-dependency 
omission score distributions for seven top-ranked words.
There are clear and large differences in how these words
impact the network's representation depending on what grammatical
function they fulfill. They all have large omission scores when they
occur as {\sc nsubj} (nominal subject) or {\sc root}, likely due to the fact that these
grammatical functions typically have a large contribution to the
complete meaning of a sentence.  Conversely, all have small omission 
scores when appearing as {\sc conj} (conjunct): this is probably because in this position
they share their contribution with the first, often more important,
member of the conjunction, for example: {\it A cow and its baby eating
  grass}. 
% The pattern for {\sc nn} (nominal modifier) is a bit more complicated: for
% four of the words shown (as well as for most other words not shown in
% the figure), the score is very low in this grammatical function--presumably 
% because most words contribute less to the sentence meaning when used  
% as modifiers than as heads (e.g.\ {\it a clock tower}). However,
% for the words {\it zebra} and {\it water}, omission scores are high
% when they act as a nominal modifier {\sc nn}. This appears due to two reasons:
% \begin{enumerate}
% \item For {\it zebra}, there are frequent erroneous parses such as {\it
%     zebra/{\sc nn} browsing/{\sc root}} instead of {\it zebra/{\sc nsubj} browsing/{\sc
%     root}}. The network does not make this mistake, and treats these
%     occurrences of {\it zebra} according to its importance as {\sc nsubj}.
% \item {\it Water} as a modifier often changes the meaning of its head in
%  a visually salient way: e.g.\ {\it water fall, water balloon, water scene, water skiing}, and
%   thus the network learns that this particular word is important in the modifier position.
% \end{enumerate}

@@

As observed in Section \ref{sec:beyondlexical}, 
adding extra information about the position of words 
explains more of the variance in the case of {\sc Visual} and especially
{\sc Textual} and {\sc LM}.
Figure \ref{fig:posrqs} shows the coefficients corresponding to the
position variables in {\sc LR~full}. Since the omission scores 
are measured at the end-of-sentence token, the expectation is that 
for {\sc Textual} and {\sc LM}, as language models,  
the words appearing closer to the end of the sentence would have a
stronger effect on the omission scores. This seems to be confirmed by
the plot as the coefficients for these two networks up until the
\emph{antepenult} are all negative.

### 2017-05-30 doc/experiments.tex

In the following section, we report a series of experiments in which
we explore the kinds of linguistic regularities the networks learn from
word-level input. Sections~\ref{sec:computeomission} to \ref{sec:beyondlexical} 
present the analyses of the final hidden activation 
vectors of the recurrent pathways from a linguistic point of view. 
Sections~\ref{sec:syntacticdim} and \ref{seq:describe} report exploratory experiments on
the linguistic features encoded by individual hidden dimensions. 
In all these experiments we report our findings based on the {\sc Imaginet}
model, whenever appropriate compare to our two other models {\sc LM} and {\sc Sum}, 
and discuss the generalizabilty of our methods to other architectures.
\label{edit:experimentsgeneral}
For all the experiments, we trained the models on the training portion of the
MSCOCO image-caption dataset \cite{lin2014microsoft}, and analyzed the
representations of the sentences in the validation set corresponding
to 5000 randomly chosen images. The target image representations were
extracted from the pre-softmax layer of the 16-layer CNN of
\namecite{simonyan2014very}.

>>>>

In this section, we report a series of experiments in which
we explore the kinds of linguistic regularities the networks learn from
word-level input. 
%Sections~\ref{sec:computeomission} to \ref{sec:beyondlexical} 
%present the analyses of the final hidden activation 
%vectors of the recurrent pathways from a linguistic point of view. 
%Sections~\ref{sec:syntacticdim} and \ref{seq:describe} report exploratory experiments on
%the linguistic features encoded by individual hidden dimensions. 
In Section~\ref{sec:computeomission} we introduce \emph{omission scores},
a metric to measure the contribution of each token to the prediction
of the networks and in Section~\ref{sec:omitimaginet} we analyze how
these scores are distributed over dependency relations and
part-of-speech categories. 
In Section~\ref{sec:beyondlexical} we investigate the extent to which
the importance of words for the different networks depend on the words themselves,
their sequential position and their grammatical function in the sentences.
\todo[inline]{add a short description of the experiments in sections 5.4 and 5.5.}

In all these experiments we report our findings based on the {\sc Imaginet}
model, whenever appropriate compare to our two other models {\sc LM} and {\sc Sum}, 
and discuss the generalizabilty of our methods to other architectures.
\label{edit:experimentsgeneral}
For all the experiments, we trained the models on the training portion of the
MSCOCO image-caption dataset \cite{lin2014microsoft}, and analyzed the
representations of the sentences in the validation set corresponding
to 5000 randomly chosen images. The target image representations were
extracted from the pre-softmax layer of the 16-layer CNN of
\namecite{simonyan2014very}.

### 2017-05-30 doc/intro.tex

We report a thorough \emph{macro level} quantitative analysis to
provide a linguistic interpretation of the networks' activation
patterns. These experiments rely on a novel method we
call \emph{omission score} to measure the importance of input tokens 
to the final prediction of models that compute distributed representations 
of sentences. This can be applied to various RNN architectures, Recursive Neural Networks 
and Convolutional Neural Networks.\label{edit:introomissiongeneral}
Based on these omission scores our experiments show that the image prediction 
pathway in general pays special attention to words of syntactic categories that 
carry semantic content. Furthermore, we observe that on top of
lexical features it learns important aspects of the information structure of sentences,
and treats the same input tokens differently depending on their
grammatical function in the sentence. In contrast, the language model
is more sensitive to the local syntactic characteristics of the input
sentences.

>>>>

\todo[inline]{once we finalise our experiments, we should update the following paragraphs and get rid of the macro/micro distinction.}
We report a thorough \emph{macro level} quantitative analysis to
provide a linguistic interpretation of the networks' activation
patterns. These experiments rely on a novel method we
call \emph{omission score} to measure the importance of input tokens 
to the final prediction of models that compute distributed representations 
of sentences. This can be applied to various RNN architectures, Recursive Neural Networks 
and Convolutional Neural Networks.\label{edit:introomissiongeneral}
Based on these omission scores our experiments show that the image prediction 
pathway in general pays special attention to words of syntactic categories that 
carry semantic content. Furthermore, we observe that on top of
lexical features it learns important aspects of the information structure of sentences,
and treats the same input tokens differently depending on their
grammatical function in the sentence. In contrast, the language model
is more sensitive to the local syntactic characteristics of the input
sentences.

### 2017-05-30 doc/models.tex

We apply our methods to the following models:
\begin{itemize}
\item {\sc Imaginet}: A multi-modal GRU network which consists of two
  pathways, {\sc Visual} and {\sc Textual}, coupled via word
  embeddings.
\item {\sc LM}: A (unimodal) language model consisting of a Gated Recurrent
  Unit (GRU) network.
\item {\sc Sum}: A network with the same objective as the {\sc Visual}
  pathway of {\sc Imaginet}, but which uses sum of word embeddings
  instead of a GRU.
\end{itemize}
The rest of this sections gives more details for these models.

@@

One of the main difficulties for training traditional Elman networks
arises from the fact that they overwrite their hidden states at every
time step with a new value computed from the current input $x_{t}$ and
the previous hidden state $\mathbf{h_{t-1}}$. Similarly to LSTMs,
Gated Recurrent Unit networks introduce a mechanism which facilitates the retention of 
information over multiple time steps.
Specifically, the GRU computes the hidden state at current time step, $\mathbf{h}_{t}$, as the
linear combination of previous activation $\mathbf{h_{t-1}}$, and a new
{\it candidate} activation $\mathbf{\tilde{h}}_t$:
%
\vspace{-.2cm}
\begin{equation}
  \mathrm{GRU}(\mathbf{h}_{t-1}, \mathbf{x}_t) = (1 - \mathbf{z}_t)\odot \mathbf{h}_{t-1} + \mathbf{z}_t \odot \mathbf{\tilde{h}}_t
\vspace{-.1cm}
\end{equation}
%
where $\odot$ is elementwise multiplication, and the update gate
activation $\mathbf{z_{t}}$ determines the amount of new information
mixed in the current state:
%
\vspace{-.1cm}
\begin{equation}
\label{eq:gru-update}
   \mathbf{z}_t = \sigma_s(\mathbf{W}_z \mathbf{x}_t + \mathbf{U}_z \mathbf{h}_{t-1})
\vspace{-.1cm}
\end{equation}
%
The candidate activation is computed as:
%
\vspace{-.2cm}
\begin{equation}
\label{eq:gru-cand}
   \mathbf{\tilde{h}}_t = \sigma(\mathbf{W} \mathbf{x}_t + \mathbf{U}(\mathbf{r}_t \odot \mathbf{h}_{t-1}))
\vspace{-.1cm}
\end{equation}
%
The reset gate $\mathbf{r_{t}}$ determines how much of the current
input $\mathbf{x_{t}}$ is mixed in the previous state
$\mathbf{h}_{t-1}$ to form the candidate activation:
%
\vspace{-.2cm}
\begin{equation}
\label{eq:gru-reset}
   \mathbf{r}_t = \sigma_s(\mathbf{W}_r \mathbf{x}_t + \mathbf{U}_r \mathbf{h}_{t-1})
\vspace{-.1cm}
\end{equation}

@@

In our version of the model we add two minor modifications 
compared to the original {\sc Imaginet}. As the performance of
{\sc Visual} is measured on image-retrieval, which is based on cosine 
distances we use cosine distance as the visual loss.
Moreover, we observe that using standardized image vectors, where each dimension 
is transformed by subtracting the mean and dividing by standard deviation further
improves performance.

@@

\subsection{Unimodal language model}
The model {\sc LM} is a language model analogous to the {\sc Textual}
pathway of {\sc Imaginet} with the difference that its word embeddings
are not shared, and its loss functions is the cross-entropy on word
prediction. Using this model we remove the visual objective as a
factor, as the model does not use the images corresponding to captions
in any way.

\subsection{Sum of word embeddings}
The model {\sc Sum} is a stripped down version of the {\sc Visual}
pathway, which does not share word embeddings, only uses the cosine
loss function, and replaces the GRU network with a summation over word
embeddings. It this removes the effect of word order from
consideration. We use this model as a baseline in the sections which
focus on language structure.

>>>>

In our analyses of the acquired linguist knowledge, we apply our methods to the following models:
\begin{itemize}
\item {\sc Imaginet}: A multi-modal Gated Recurrent Unit (GRU) network consisting of two
  pathways, {\sc Visual} and {\sc Textual}, coupled via word embeddings.
\item {\sc LM}: A (unimodal) language model consisting of a GRU network.
\item {\sc Sum}: A network with the same objective as the {\sc Visual}
  pathway of {\sc Imaginet}, but which uses sum of word embeddings
  instead of a GRU.
\end{itemize}
The rest of this section gives a detailed description of these models.

@@

One of the main difficulties for training traditional Elman networks
arises from the fact that they overwrite their hidden states at every
time step with a new value computed from the current input $x_{t}$ and
the previous hidden state $\mathbf{h_{t-1}}$. Similarly to LSTMs,
Gated Recurrent Unit networks introduce a mechanism which facilitates the retention of 
information over multiple time steps.
Specifically, the GRU computes the hidden state at current time step $\mathbf{h}_{t}$, as the
linear combination of previous activation $\mathbf{h_{t-1}}$, and a new
{\it candidate} activation $\mathbf{\tilde{h}}_t$:
%
\vspace{-.2cm}
\begin{equation}
  \mathrm{GRU}(\mathbf{h}_{t-1}, \mathbf{x}_t) = (1 - \mathbf{z}_t)\odot \mathbf{h}_{t-1} + \mathbf{z}_t \odot \mathbf{\tilde{h}}_t
\vspace{-.1cm}
\end{equation}
%
where $\odot$ is elementwise multiplication, and the update gate
activation $\mathbf{z_{t}}$ determines the amount of new information
mixed in the current state:
%
\vspace{-.1cm}
\begin{equation}
\label{eq:gru-update}
   \mathbf{z}_t = \sigma_s(\mathbf{W}_z \mathbf{x}_t + \mathbf{U}_z \mathbf{h}_{t-1})
\vspace{-.1cm}
\end{equation}
%
The candidate activation is computed as:
%
\vspace{-.2cm}
\begin{equation}
\label{eq:gru-cand}
   \mathbf{\tilde{h}}_t = \sigma(\mathbf{W} \mathbf{x}_t + \mathbf{U}(\mathbf{r}_t \odot \mathbf{h}_{t-1}))
\vspace{-.1cm}
\end{equation}
%
The reset gate $\mathbf{r_{t}}$ determines how much of the current
input $\mathbf{x_{t}}$ is mixed in the previous state
$\mathbf{h}_{t-1}$ to form the candidate activation:
%
\vspace{-.2cm}
\begin{equation}
\label{eq:gru-reset}
   \mathbf{r}_t = \sigma_s(\mathbf{W}_r \mathbf{x}_t + \mathbf{U}_r \mathbf{h}_{t-1})
\vspace{-.1cm}
\end{equation}

@@

In our version of the model we add two minor modifications 
compared to the original {\sc Imaginet}. As the performance of
{\sc Visual} is measured on image-retrieval which is based on cosine 
distances, we use cosine distance as the visual loss.
Moreover, we observe that using standardized image vectors, where each dimension 
is transformed by subtracting the mean and dividing by standard deviation, further
improves performance.

@@

\subsection{Unimodal language model}
The model {\sc LM} is a language model analogous to the {\sc Textual}
pathway of {\sc Imaginet} with the difference that its word embeddings
are not shared, and its loss function is the cross-entropy on word
prediction. Using this model we remove the visual objective as a
factor, as the model does not use the images corresponding to captions
in any way.

\subsection{Sum of word embeddings}
The model {\sc Sum} is a stripped-down version of the {\sc Visual}
pathway, which does not share word embeddings, only uses the cosine
loss function, and replaces the GRU network with a summation over word
embeddings. This removes the effect of word order from
consideration. We use this model as a baseline in the sections which
focus on language structure.
